Joint Likelihood Deconvolution of Astronomical Images in the Presence of Poisson Noise

### 2023-10-26 src/tex/ms.tex
[showyourwork] overleaf sync

    \begin{itemize}
        \item Results are available in \url{https://github.com/jolideco/jolideco-benchmarks}
        \item We tried the binary search tree as proposed in \cite{Parameswaran2018},
        but without success, most likely because the native \texttt{Pytorch} support
        for parallelization speeds up the standard GMM quite a bit...
        \item Could use a Neural Net for GMM component selection as proposed in \cite{Rosenbaum15}
        \item GPU support via Pytorch
    \end{itemize}

@@

    \begin{table}
        \centering
        \begin{tabular}{ |c|c|c| } 
        \input{output/chandra-obs-table.tex}
        \end{tabular}
        \label{tab:chandra_obs}
        \script{chandra-obs-table.py}
    \end{table}

@@

    \vlk{We compute the full-width at half-max (FWHM) of each of the profiles, for both the original counts and the deconvolved intensities, and find that the deconvolved profile is narrower.  For the counts profile, FWHM$_C=0.63 \pm 0.09$~arcsec, where the error bar is obtained by computing the standard deviations over the bootstrap sample; and the deconvolved intensity profile has FWHM$_J=0.56 \pm 0.03$~arcsec.  Note that the PSF itself has a FWHM$_{{\rm PSF}\approx{0.6}}$~arcsec\footnote{See projected radial profiles of the on-axis PSF at \url{https://space.mit.edu/cxc/marx/tests/PSF.html}.  The 1$\sigma$ radius is $\approx$0.26~arcsec (corresponding to 39\% enclosed counts in 2-D), suggesting that FWHM$_{{\rm PSF}}\approx{2.355{\cdot}0.26=0.62}$~arcsec.}.  }
    The profile computed from the images reconstructed with \jolideco has a much smaller noise level and the peak is much more pronounced. By fitting the profile with a univariate spline and finding its roots we measure the full width half maximum of the curves. We find $\lambda_J=0.56\pm0.03$ for the \jolideco profile and $\lambda_C=0.63\pm0.09$ for the profile computed on the counts image. Assuming that the PSF and the shape of peak of the profile are approximately Gaussian a we expect the following relation between the width of the counts profile, the PSF and the flux profile:

    \begin{equation}
        \lambda_{J} = \sqrt{\lambda_C^2 - \lambda_{\mathrm{PSF}}^2}
    \end{equation}

    With $\lambda_{\mathrm{PSF}} \approx 0.25 \mathrm{arcsec}$ \footnote{The 38\% containment radius of the Chandra PSF is $\sigma_{39}=2$ native pixels. This corresponds to  see e.g. \url{https://space.mit.edu/cxc/marx/tests/PSF.html}}
    we obtain ..., which is consistent with the values we measured for the profile. The profile computed from the images reconstructed with \jolideco show both a smaller width of the profile as well as a much smaller error.

@@

    The \texttt{fermitools} produce an energy dependent counts, exposure and PSF cube. To further reduce the data to images we use the \textit{Gammapy} package v1.0 \citep{GammapyZenodov1.0.1, Gammapy2023}. To compute an image of the expected background emission we rely on the latest Galactic diffuse emission model \texttt{gll\_iem\_v07.fits} and also include nearby sources from the 3FHL catalog \citep{Ajello2017}. We first fit the model components to the data using \textit{Gammapy}, excluding a circular region of radius \qty[mode = text]{1.02}{$\deg$} centered on Vela Junior. Finally we reduce this model to an image by summing over the energy dimension. To compute the exposure and PSF image we assume a power law with index $\Gamma=1.77$ \citep{Ajello2017} as a spectral model for Vela Junior and integrate over the energy dimension of the cube while weighting with the spectral model. We chose the GMM patch prior learned from \chandra SNRs described in the Section~\ref{sssec:chandra-snrs}. Because of the limited statistics we did not use any oversampling but keep the pixel size of the counts image.

    Figure~\ref{fig:example-fermi-lat} shows the data, result and residuals. The \fermi data is sparse, with very limited statistics. \jolideco clearly improves the noise level as well as angular resolution of the data. The calibration of the predicted counts finds different background model norms and positional shifts of order of $<1$ pixel for the different event classes. This way it corrects for systematic errors in the reconstruction of the position. One can see that the reconstruction includes background emission around Vela Junior, indicating that the Galactic diffuse background model is either underestimated in brightness and / or associated with systematic errors. The background estimate could be improved by choosing a larger field of view. However this would also require a larger effort of modeling nearby sources, which we avoided here.

    To further check the quality of the model we compute a map of approximate residual significance defined by:

@@

    Where $\hat{N}$ is the weighted sum of counts from a Gaussian smoothing kernel with $\sigma=5$ pixels. The result is shown in the right panel of Fig.~\ref{fig:example-fermi-lat}. The image
    shows no systematic effects and a spatially homogeneous distribution of the residual noise. The resulting distribution of residual significances per pixel is centered at $\mu=-0.001$ with a width of $\sigma=0.991$, indicating a close agreement with a unit Gaussian distribution, as expected for a good model of the emission. More details of the \fermi analysis are available at \url{https://github.com/jolideco/jolideco-fermi-examples}

@@

    \appendix
    \section{Source Patterns}

@@

>>>>

    \begin{itemize}
        \item Results are available in \url{https://github.com/jolideco/jolideco-benchmarks}
        \item We tried the binary search tree as proposed in \cite{Parameswaran2018},
        but without success, most likely because the native \texttt{Pytorch} support
        for parallelization speeds up the standard GMM quite a bit...
        \item GPU support via Pytorch
    \end{itemize}

@@

@@

    \vlk{We compute the full-width at half-max (FWHM) of each of the profiles, for both the original counts and the deconvolved intensities, and find that the deconvolved profile is narrower.  For the counts profile, FWHM$_C=0.63 \pm 0.09$~arcsec, where the error bar is obtained by computing the standard deviations over the bootstrap sample; and the deconvolved intensity profile has FWHM$_J=0.56 \pm 0.03$~arcsec.  Note that the PSF itself has a FWHM$_{{\rm PSF}\approx{0.6}}$~arcsec\footnote{See projected radial profiles of the on-axis PSF at \url{https://space.mit.edu/cxc/marx/tests/PSF.html}.  The 1$\sigma$ radius is $\approx$0.26~arcsec (corresponding to 39\% enclosed counts in 2-D), suggesting that FWHM$_{{\rm PSF}}\approx{2.355{\cdot}0.26=0.62}$~arcsec.}.  }
    The profile computed from the images reconstructed with \jolideco has a much smaller noise level and the peak is much more pronounced. By fitting the profile with a univariate spline and finding its roots we measure the full width half maximum of the curves. We find $w_J=0.56\pm0.03$ for the \jolideco profile and $w_C=0.63\pm0.09$ for the profile computed on the counts image. Assuming that the PSF and the shape of peak of the profile are approximately Gaussian a we expect the following relation between the width of the counts profile, the PSF and the flux profile:

    \begin{equation}
        w_{J} = \sqrt{w_C^2 - \w_{\mathrm{PSF}}^2}
    \end{equation}

    With $w_{\mathrm{PSF}} \approx 0.25 \mathrm{arcsec}$ \footnote{The 38\% containment radius of the Chandra PSF is $\sigma_{39}=2$ native pixels. This corresponds to  ee e.g. \url{https://space.mit.edu/cxc/marx/tests/PSF.html}}
    we obtain ..., which is consistent with the values we measured for the profile. The profile computed from the images reconstructed with \jolideco show both a smaller width of the profile as well as a much smaller error.

@@

    The \texttt{fermitools} produce an energy dependent counts, exposure and PSF cube. To further reduce the data to images we use the \textit{Gammapy} package v1.0 \citep{GammapyZenodov1.0.1, Gammapy2023}. To compute an image of the expected background emission we rely on the latest Galactic diffuse emission model \texttt{gll\_iem\_v07.fits} and also include nearby sources from the 3FHL catalog \citep{Ajello2017}. We first fit the model components to the data using \textit{Gammapy}, excluding a circular region of radius \qty[mode = text]{1.02}{$\deg$} centered on Vela Junior. Finally we reduce this model to an image by summing over the energy dimension. To compute the exposure and PSF image we assume a power law with index $\Gamma=1.77$ \citep{Ajello2017} as a spectral model for Vela Junior and integrate over the energy dimension of the cube while weighting with the spectral model.

    We chose the GMM patch prior learned from \chandra SNRs described in the Section~\ref{sssec:chandra-snrs}. Because of the limited statistics we did not use any oversampling but keep the pixel size of the counts image. 
    Figure~\ref{fig:example-fermi-lat} shows the data, result of the \jolideco method and corresponding residuals. The \fermi data is sparse, with very limited number of signal counts. \jolideco clearly improves the noise level as well as angular resolution of the data. The calibration parameters of the predicted counts converged to different norms for the background model and positional shifts of order of $<1$ pixel for the different event classes. This way \jolideco corrects for systematic errors in the reconstruction of the position. One can see that the reconstruction includes background emission around Vela Junior, indicating that the Galactic diffuse background model is either underestimated in brightness and / or associated with systematic errors. The background estimate could be improved by choosing a larger field of view. However this would also require a larger effort of modeling nearby sources, which we avoided here.

    To further check the quality of the reconstructed flux we compute a map of approximate residual significance defined by:

@@

    Where $\hat{N}$ is the weighted sum of counts from a Gaussian smoothing kernel with $\sigma=5$ pixels. The result is shown in the right panel of Fig.~\ref{fig:example-fermi-lat}. The image
    shows no systematic effects and a spatially homogeneous distribution of the residual noise. The resulting distribution of residual significances per pixel is centered at $\mu=-0.001$ with a width of $\sigma=0.991$, indicating a close agreement with a standard normal distribution, as expected for a good model of the emission. More details of the \fermi analysis are available at \url{https://github.com/jolideco/jolideco-fermi-examples}

@@

    \appendix
    \section{Details on Source Patterns}

@@

    \section{Chandra Observation Table}

    \begin{table}
    \centering
        \begin{tabular}{ c|c|c|c } 
            \input{output/chandra-obs-table.tex}
        \end{tabular}
        \label{tab:chandra_obs}
        \script{chandra-obs-table.py}
        \caption{Observation list and details from the Chandra example analysis in Section~\ref{sec:chandra-example}}
    \end{table}
